---
title: "Citation styles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citation styles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Which styles are bundled

`uomthesis` ships five CSL (Citation Style Language) files that cover the most common citation styles used in AMBS research. View them with `list_csl()`:

```{r list-csl}
library(uomthesis)
list_csl()
```

Each style is identified by a short `name` used in `create_thesis()` and `copy_csl()`. The `sha256` column records the hash of the CSL file at retrieval time so that reproducibility can be verified: if the upstream file changes, the hash will differ.

The bundled styles are:

| Name | Full style name | Common in |
|---|---|---|
| `harvard-manchester` | Harvard (Cite Them Right) | Business, social sciences |
| `apa` | APA 7th edition | Psychology, education |
| `chicago-author-date` | Chicago author-date | History, humanities |
| `mhra` | MHRA footnotes | Humanities, languages |
| `vancouver` | Vancouver (Elsevier variant) | Medicine, life sciences |

Policy section 7.2 requires citations to be in a consistent, recognised style throughout the thesis. Any of the five bundled styles satisfies this requirement.

## Picking one at scaffold time

Pass `reference_style` to `create_thesis()`:

```{r scaffold-style, eval = FALSE}
create_thesis(
  "my-thesis",
  degree          = "PhD",
  author          = list(forename = "Jane", surname = "Doe"),
  title           = "An analysis of something",
  year            = 2027,
  division        = "MSM",
  reference_style = "apa"
)
```

The scaffolder copies `apa.csl` into the extension's `csl/` subdirectory and sets `csl: _extensions/uomthesis-standard/csl/apa.csl` in `_quarto.yml`.

If `reference_style` is omitted the default is `harvard-manchester`.

## Switching styles after scaffolding

To change the citation style in an existing project:

**Step 1.** Copy the new CSL into your extension directory:

```{r copy-csl, eval = FALSE}
copy_csl("apa", to = "_extensions/uomthesis-standard/csl/")
```

**Step 2.** Update the `csl:` line in `_quarto.yml`:

```yaml
bibliography: references.bib
csl: _extensions/uomthesis-standard/csl/apa.csl
```

**Step 3.** Re-render:

```{r re-render, eval = FALSE}
quarto::quarto_render()
```

For a journal-format project, replace `uomthesis-standard` with `uomthesis-journal` in the path.

## Bringing your own CSL

You are not restricted to the five bundled styles. If your supervisor requires a specific style, or if your target journal has a custom CSL:

1. Download the CSL file (e.g., from <https://citationstyles.org/> or the CSL repository at <https://github.com/citation-style-language/styles>).
2. Save it anywhere inside your project (e.g., `_extensions/uomthesis-standard/csl/my-style.csl`).
3. Point `csl:` in `_quarto.yml` at the file path:

```yaml
csl: _extensions/uomthesis-standard/csl/my-style.csl
```

The `csl-bundled-or-path-exists` validator rule accepts any path that resolves to an existing file, so your custom CSL will pass validation as long as it is present on disk.

## How bundled styles are pinned

The file `inst/csl/SOURCES.yml` records the provenance of each bundled CSL:

```yaml
- name: harvard-manchester
  url: https://raw.githubusercontent.com/.../harvard-cite-them-right.csl
  retrieved: '2026-06-02'
  sha256: 0ca3d2c7...
```

When a new version of the package is built, `dev/sync_shared.R` re-downloads each file, verifies the sha256, and updates `SOURCES.yml` if the upstream file has changed. This means you can always tell exactly which version of a CSL style your thesis used and reproduce it from the hash. The `list_csl()` function surfaces this information directly:

```{r sources, eval = FALSE}
# View provenance of all bundled styles
list_csl()[, c("name", "retrieved", "sha256")]
```
