---
title: "Compliance validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compliance validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`check_thesis()` runs a registry of source-phase validation rules against your Quarto thesis project. Each rule maps to one or more sections of the *Presentation of Theses Policy* (v12, March 2026). A finding is either an **error** (the thesis will likely be rejected unless fixed) or a **warning** (should be investigated; may be intentional).

Source-phase rules inspect `.qmd` files, `_quarto.yml`, and the extension partials. A planned PDF-phase rule set (v0.2) will additionally inspect the rendered document.

Run the validator from within your thesis project directory:

```{r check-basic, eval = FALSE}
library(uomthesis)
check_thesis()
```

## The rule registry

The table below lists all 19 rules in registry order.

```{r rule-table, echo = FALSE}
# Build a summary table from the live registry (runs at vignette build time
# only -- readers who do not execute the code see the pre-rendered table).
if (requireNamespace("uomthesis", quietly = TRUE)) {
  registry <- uomthesis:::rule_registry()
  tbl <- data.frame(
    ID           = vapply(registry, `[[`, character(1), "id"),
    `Policy ref` = vapply(registry, `[[`, character(1), "policy_ref"),
    Severity     = vapply(registry, `[[`, character(1), "severity"),
    Formats      = vapply(registry, function(r) {
      f <- r$formats %||% "both"
      if (is.null(f) || identical(f, c("standard", "journal"))) "both" else f
    }, character(1)),
    check.names  = FALSE,
    stringsAsFactors = FALSE
  )
  knitr::kable(tbl)
}
```

### metadata-complete

**Policy section:** 8 / general submission requirements
**Severity:** error
**Formats:** both

Checks that all required `uomthesis:` YAML fields are present and non-empty in `index.qmd`: `degree`, `author` (with `forename` and `surname`), `title`, `year`, `division`, `faculty`, `school`, and `thesis_format`.

**How to fix:** Open `index.qmd` and ensure every field in the `uomthesis:` block has a non-empty value. Use `validate_metadata()` for a focused check on just the metadata block.

---

### degree-faculty-school

**Policy section:** general submission
**Severity:** error
**Formats:** both

Checks that `degree`, `faculty`, and `school` each belong to the allowed sets. Supported degrees: PhD, MPhil, MD, EdD, DBA, EngD. Supported faculty: Humanities. Supported school: Alliance Manchester Business School.

**How to fix:** Verify the values in your `uomthesis:` block against the allowed sets. If your degree is not listed, open a GitHub issue -- the registry is updated when the Doctoral Academy recognises new degree types.

---

### year-not-month

**Policy section:** general submission
**Severity:** error
**Formats:** both

Checks that `year` is a plausible four-digit integer (2000--2100) and not a month name or other non-year string.

**How to fix:** Set `year` to a numeric year such as `2027`, not a string like `"March 2027"`.

---

### thesis-format

**Policy section:** 13
**Severity:** error
**Formats:** both

Checks that `thesis_format` is either `"standard"` or `"journal"`.

**How to fix:** Set `thesis_format` to `standard` or `journal` in the `uomthesis:` block.

---

### font-allowed

**Policy section:** 7.1
**Severity:** error
**Formats:** both

Checks that `mainfont` is one of the policy-permitted typefaces. Run `policy_info()` to see the current permitted list.

**How to fix:** Choose a permitted font. `TeX Gyre Termes` (a Times New Roman equivalent) is the recommended default and is available in all standard LaTeX distributions.

---

### font-engine-compat

**Policy section:** 7.1
**Severity:** error
**Formats:** both

Checks that the chosen `mainfont` is compatible with the configured PDF engine. Some OpenType fonts require LuaLaTeX or XeLaTeX and cannot be used with pdflatex.

**How to fix:** Either switch to LuaLaTeX (recommended) by setting `pdf-engine: lualatex` in `_quarto.yml`, or choose a font that is pdflatex-safe.

---

### linespacing-allowed

**Policy section:** 7.1
**Severity:** error
**Formats:** both

Checks that `linestretch` is 1.5 or 2.0 (the only values the policy permits for main text). The field may be absent (the extension default of 1.5 will be used).

**How to fix:** Set `linestretch` to `1.5` or `2.0`.

---

### ai-disclosure-shape

**Policy section:** 9.1.d
**Severity:** warning
**Formats:** both

If `ai_disclosure.include` is `true`, checks that `ai_disclosure.tools` is a non-empty list of tool names. A declaration without a tools list is malformed.

**How to fix:** Either set `ai_disclosure.include: false` (if you did not use AI tools) or populate `ai_disclosure.tools` with the names of each tool used (e.g., `[ChatGPT-4o, GitHub Copilot]`).

---

### english-language

**Policy section:** 6.1
**Severity:** warning
**Formats:** both

If the `lang` key is set in `_quarto.yml` or `index.qmd`, checks that it starts with `en`. Policy section 6.1 requires the thesis to be written in UK English unless advance written approval has been obtained for another language.

**How to fix:** If the warning is incorrect (you have approval for another language), suppress this rule by passing `rules = setdiff(uomthesis:::rule_ids(), "english-language")` to `check_thesis()`. Otherwise set `lang: en-GB`.

---

### title-page-statement

**Policy section:** 8.1.b
**Severity:** error
**Formats:** both

Checks that the title page partial (`_extensions/uomthesis-*/partials/title-page.tex`) contains the policy-mandated submission statement. Run `policy_info()` to see the required text.

**How to fix:** Do not edit the title-page partial directly unless you know what you are doing. If you have accidentally altered it, re-copy it from the package: `uomthesis:::install_partial("title-page.tex", format = "standard")`.

---

### declaration-text

**Policy section:** 8.1.f
**Severity:** error
**Formats:** both

Checks that the declaration partial (`_extensions/uomthesis-*/partials/declaration.tex`) contains one of the two policy-mandated declaration texts verbatim. Run `policy_info()` to see both variants.

**How to fix:** Do not paraphrase the declaration. Restore it from the package partial if it has been altered.

---

### copyright-text

**Policy section:** 8.1.g
**Severity:** error
**Formats:** both

Checks that the copyright partial contains all four mandatory bullet points that policy section 8.1.g requires. Run `policy_info()` to see the required text.

**How to fix:** Restore the copyright partial from the package template. Do not remove or alter the mandatory bullets.

---

### copyright-author-match

**Policy section:** 8.1.g
**Severity:** error
**Formats:** both

Checks that the author name in the copyright partial matches the `author` metadata in `index.qmd`. A mismatch typically means the copyright partial was copied from another candidate's project without being updated.

**How to fix:** Search the copyright partial for the previous author's name and replace it with your own name, or re-run `create_thesis()` in a fresh directory.

---

### prelim-order

**Policy section:** 8.1
**Severity:** error
**Formats:** both

Checks that the preliminary pages appear in the order mandated by policy section 8.1 (title, abstract, declaration, copyright, dedication/acknowledgements if present, then body chapters, then appendices).

**How to fix:** Reorder the `chapters:` list in `_quarto.yml` so that preliminary files appear before body files. The validator reports the first out-of-order file it finds.

---

### abstract-one-page

**Policy section:** 8.1.e
**Severity:** warning
**Formats:** both

Checks (heuristically) whether the abstract in `index.qmd` exceeds approximately one page of text. The source-phase check counts words; the PDF-phase check (v0.2) will verify the rendered page count.

**How to fix:** Shorten your abstract to no more than ~300--350 words. The policy mandates one page maximum.

---

### csl-bundled-or-path-exists

**Policy section:** 7.2
**Severity:** error
**Formats:** both

Checks that the `csl:` value in `_quarto.yml` is either a bundled style name (resolvable via `list_csl()`) or a path to a file that exists on disk.

**How to fix:** Use one of the bundled style names (run `list_csl()$name` to see them), or provide a valid path to a CSL file. See `vignette("citation-styles")`.

---

### bibliography-exists

**Policy section:** 7.2
**Severity:** error
**Formats:** both

Checks that every file listed under `bibliography:` in `_quarto.yml` exists on disk.

**How to fix:** Create the missing `.bib` file, or remove it from the `bibliography:` list in `_quarto.yml`.

---

### journal-rationale-present

**Policy section:** 13.10
**Severity:** error
**Formats:** journal only

Checks that a chapter file whose name contains `rationale` (case-insensitive) is listed in `_quarto.yml` under `book.chapters`.

**How to fix:** Add a rationale chapter (e.g., `chapters/00-rationale.qmd`) and list it in `_quarto.yml`. See `vignette("journal-format")`.

---

### journal-contribution-stmts

**Policy section:** 13.3
**Severity:** warning
**Formats:** journal only

Checks that each chapter file whose name contains `paper` (case-insensitive) carries a `contribution` attribute on its top-level heading.

**How to fix:** Add `{contribution="..."}` to the `#` heading in each paper chapter. See `vignette("journal-format")` for the expected format.

---

## Output formats

`check_thesis()` supports three output formats via the `format` argument:

### Console (default)

Findings are printed to the console using `cli` styling. Errors are highlighted in red, warnings in yellow. A summary line reports the total counts.

```{r console, eval = FALSE}
check_thesis()
```

### Markdown

Findings are written to `check-report.md` in the project root. The file is overwritten on each run. Useful for committing a compliance snapshot to version control.

```{r markdown, eval = FALSE}
check_thesis(".", format = "markdown")
# Writes check-report.md
```

### JSON

Returns a JSON string (invisibly) and also prints it. Each finding is a JSON object with fields: `rule_id`, `severity`, `message`, `location`, `policy_ref`, `hint`. Useful for CI integration.

```{r json, eval = FALSE}
result <- check_thesis(".", format = "json")
# result is a JSON string
```

## Fail-on behaviours

The `fail_on` argument controls when `check_thesis()` signals an R error (allowing use in CI pipelines):

| Value | Behaviour |
|---|---|
| `"none"` (default) | Never signals an error; always returns invisibly |
| `"warning"` | Signals an error if any warning or error findings are found |
| `"error"` | Signals an error only if error-severity findings are found |

```{r fail-on, eval = FALSE}
# Stop the build if any errors are found:
check_thesis(".", fail_on = "error")

# Stop the build if any warnings or errors are found:
check_thesis(".", fail_on = "warning")
```

## Restricting which rules run

Pass a character vector of rule IDs to `rules` to run only a subset:

```{r rules-subset, eval = FALSE}
check_thesis(".", rules = c("font-allowed", "linespacing-allowed"))
```

To get the full list of rule IDs programmatically:

```{r rule-ids, eval = FALSE}
vapply(uomthesis:::rule_registry(), `[[`, character(1), "id")
```
