---
title: "Journal-format theses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Journal-format theses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## When journal format applies

Policy section 13 permits PhD candidates to submit a thesis as a collection of papers rather than a conventional monograph. The *Journal Format Theses Guiding Principles* (The University of Manchester) apply. Key points:

- The thesis must still be a coherent, unified piece of work with an overarching argument.
- There is no minimum or maximum number of papers, but the overall word count cap is **90,000 words** for a PhD journal-format thesis (versus 80,000 for standard format).
- Papers may be published, submitted, or in preparation. Published or submitted papers do not need to be reformatted to match the body font, but must be incorporated as chapters.
- Each paper chapter must carry an explicit statement of the candidate's contribution (policy section 13.3).
- A dedicated **rationale chapter** is mandatory (policy section 13.10).

Approval to use journal format must be obtained from the School before the thesis is submitted. Contact the Doctoral Academy for the approval form.

## Scaffolding a journal project

Pass `format = "journal"` to `create_thesis()`:

```{r scaffold-journal, eval = FALSE}
library(uomthesis)

create_thesis(
  "my-journal-thesis",
  degree  = "PhD",
  author  = list(forename       = "Jane",
                 middle_initial = "Q",
                 surname        = "Doe"),
  title   = "Three papers on important things",
  year    = 2027,
  division = "MSM",
  format  = "journal"
)
```

The scaffolded project includes:

- `chapters/00-rationale.qmd` -- the mandatory rationale chapter (pre-populated with section headings).
- `chapters/01-paper-one.qmd`, `chapters/02-paper-two.qmd` -- starter paper chapters with contribution declaration attributes.
- `_extensions/uomthesis-journal/` -- the journal-format Quarto extension.

## The rationale chapter

Policy section 13.10 mandates a rationale chapter that:

1. Explains why the journal format is appropriate for this thesis.
2. Describes how the papers relate to each other and to the overarching argument.
3. States the role of each paper in the thesis structure.

The scaffolded `00-rationale.qmd` contains placeholder headings for each of these points. Fill them in before you write anything else -- they will anchor your writing throughout.

The validator rule `journal-rationale-present` checks that a file whose name contains `rationale` (case-insensitive) exists among the book chapters. If you rename the file, update `_quarto.yml` accordingly.

## Per-chapter contribution declarations

Policy section 13.3 requires that each paper chapter contains an explicit statement of the candidate's individual contribution. In `uomthesis`, the journal extension reads a `contribution` attribute on the chapter's top-level heading.

In your `.qmd` file, add the attribute to the `#` heading:

```markdown
# Paper title {contribution="Lead author. Conceived and conducted all analyses, wrote the first draft. Co-author X provided comments on three manuscript revisions."}
```

The extension renders this into a formatted contribution box at the start of the chapter. The text should be specific: name each co-author, describe their role, and state clearly what the candidate did independently.

The validator rule `journal-contribution-stmts` checks for the presence of a `contribution` attribute (or equivalent YAML key) in each file whose name contains `paper` (case-insensitive). If a file is missing the attribute it is reported as a warning.

## Word cap

For a PhD journal-format thesis the cap is **90,000 words** of main text. The same exclusions apply as for standard format (preliminary pages, bibliography, appendices, footnotes).

```{r wc-journal, eval = FALSE}
# Run from within the thesis project:
word_count()
```

`word_count()` reads the thesis format from your `uomthesis:` YAML block and applies the correct cap automatically.

## Validator differences

When your `uomthesis.thesis_format` is set to `"journal"`, `check_thesis()` runs two additional rules:

| Rule ID | Policy section | What it checks |
|---|---|---|
| `journal-rationale-present` | 13.10 | A chapter file named `*rationale*` exists in `_quarto.yml` |
| `journal-contribution-stmts` | 13.3 | Each `*paper*` chapter carries a `contribution` attribute |

Both rules are skipped entirely for standard-format theses.

Run the validator with:

```{r check-journal, eval = FALSE}
check_thesis()
```

If you want to run only the journal-specific rules:

```{r check-journal-rules, eval = FALSE}
check_thesis(".", rules = c("journal-rationale-present",
                             "journal-contribution-stmts"))
```

See `vignette("compliance")` for the complete rule list and output format options.
